\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-24 papers/jfp/3-abstractions.tex
Finish S3.8

We really, really want \hs{Task} to be \emph{polymorphic} in \hs{f}.
Given \emph{one} task description \cmd{T}, we want to explore \emph{many} build
systems that can build \cmd{T} -- and we will do so in sections~\S\ref{sec-build}
and~\S\ref{sec-implementations}. As we shall see, each build system will use a
different \hs{f}, so the task description must not fix \hs{f}.

@@

\subsection{Computing Dependencies}\label{sec-deps}\label{secdeps}

@@

\subsection{Examples of Tasks}\label{sub-task-examples}

The \hs{Task} abstraction can be used to express pure functions in a way that is
convenient for their memoization, as will be further discussed
in~\S\ref{sec-related-memo}. Below we give examples of tasks corresponding to
several well-known functions.

The \emph{Fibonacci sequence} $F_n = F_{n-1} + F_{n-2}$ is the one of the most
favourite examples in the functional programming community and so we start with
it:

@@

\noindent
Here the keys $n < 2$ are input parameters, and one can obtain the usual
Fibonacci sequence by picking 0 and 1 for $n = 0$ and $n = 1$, respectively.
Any minimal build system will compute the sequence with memoization, i.e.
without recomputing the same value twice.

% TODO: Add a figure with a dependency graph?
The dependencies of Fibonacci numbers are statically known, hence we can express
it using \hs{Tasks}~\hs{Applicative}, benefiting from static dependency
analysis:

@@

\noindent
Interestingly, the \emph{Ackermann function} -- a famous example of a function
that is not primitive recursive -- cannot be expressed as an applicative task,
since it needs to perform an intermediate recursive call to determine one of its
dependencies $A(m, n-1)$:

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
ackermann :: Tasks Monad (Integer, Integer) Integer
ackermann (n, m)
    | m < 0 || n < 0 = Nothing
    | m == 0    = Just $ Task $ const $ pure (n + 1)
    | n == 0    = Just $ Task $ \fetch -> fetch (m - 1, 1)
    | otherwise = Just $ Task $ \fetch -> do index <- fetch (m, n - 1)
                                             fetch (m - 1, index)
\end{minted}
\vspace{1mm}

\noindent
Now that we have seen examples of applicative and monadic tasks, let us finish
with an example of a functorial task -- the \emph{Collatz sequence}, where
given an initial value $c_0$, we calculate the next value $c_n$ from $c_{n-1}$
either by dividing $c_{n-1}$ by 2 (if it is even) or multiplying it by 3 and
adding 1 (if it is odd):

@@

Note that functorial tasks allow us to statically extract even more precise
information about their dependencies than applicative tasks. Indeed, we
statically know that a \hs{Task}~\hs{Functor} has exactly one \hs{dependency}:

@@

>>>>

We really, really want \hs{Task} to be \emph{polymorphic} in \hs{f}.
Given \emph{one} task description \cmd{T}, we want to explore \emph{many} build
systems that can build \cmd{T} -- and we will do so in
section~\S\ref{sec-implementations}. As we shall see, each build system will use
a different \hs{f}, so the task description must not fix \hs{f}.

@@

\subsection{Computing Dependencies}\label{sec-deps}

@@

\subsection{Examples of Tasks}\label{sec-task-examples}

In this section we give examples of tasks whose definitions involve different
constraints on the computation context: \hs{Functor}, \hs{Applicative},
\hs{Monad} and \hs{MonadState}~\hs{s}.

We start with one of the favourite examples for functional
programmers -- the \emph{Fibonacci sequence} $F_n = F_{n-1} + F_{n-2}$:

@@

\noindent
Here the keys $n < 2$ are input parameters, and one can obtain the usual
Fibonacci sequence by picking $F_0=0$ and $F_1=1$, respectively. Any minimal
build system will compute the sequence with memoization, i.e. without
recomputing the same value twice.

% TODO: Add a figure with a dependency graph?
Dependencies of elements of the Fibonacci sequence are known statically, hence
we can express it using \hs{Tasks}~\hs{Applicative}, and benefit from static
dependency analysis (\S\ref{sec-deps}):

@@

\noindent
Interestingly, the \emph{Ackermann function} -- a famous example of a function
that is not primitive recursive -- cannot be expressed as an applicative task,
because it needs to perform an intermediate recursive call to determine the
value of one of its dependencies $A(m, n-1)$:

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
ackermann :: Tasks Monad (Integer, Integer) Integer
ackermann (m, n)
    | m < 0 || n < 0 = Nothing
    | m == 0    = Just $ Task $ const $ pure (n + 1)
    | n == 0    = Just $ Task $ \fetch -> fetch (m - 1, 1)
    | otherwise = Just $ Task $ \fetch -> do index <- fetch (m, n - 1)
                                             fetch (m - 1, index)
\end{minted}
\vspace{1mm}

\noindent
Functorial tasks are less common than applicative and monadic, but there is a
classic example too -- the \emph{Collatz sequence}, where given an initial value
$c_0$, we calculate the next value $c_n$ from $c_{n-1}$ either by dividing
$c_{n-1}$ by 2 (if it is even) or multiplying it by 3 and adding 1 (if it is
odd):

@@

Note that we can statically extract even more precise dependency information
from functorial tasks compared to applicative tasks. Indeed, we statically know
that a \hs{Task}~\hs{Functor} has \emph{exactly one} \hs{dependency}:

@@

\noindent
The \hs{Tasks} abstraction allows us to express pure functions in a way that
is convenient for their memoization and incremental recomputation
(see~\S\ref{sec-related-memo} for a discussion on memoization). If we
furthermore need to share computation results via a cloud cache, we can use
\hs{Tasks}~\hs{(MonadState}~\hs{s)} that will
play an important role in~\S\ref{sec-implementations}. Intuitively, by making
a shared state of type \hs{s} available to a task, we give it the abilities to
lookup and store computation results in a cache by using the \hs{MonadState}
methods \hs{get} and \hs{set}. For example, below we implement a cloud version
of the Ackermann task that uses a \hs{Cache} of type
\hs{Map}~\hs{(Integer,}~\hs{Integer)}~\hs{Integer} for sharing results of known
Ackermann values.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Cache = Map (Integer, Integer) Integer

cloudAckermann :: Tasks (MonadState Cache) (Integer, Integer) Integer
cloudAckermann (m, n)
    | m < 0 || n < 0 = Nothing
    | m == 0    = Just $ Task $ const $ pure (n + 1)
    | n == 0    = Just $ Task $ \fetch -> fetch (m - 1, 1)
    | otherwise = Just $ Task $ \fetch -> do
        cache <- get
        case Map.lookup (m, n) cache of
            Nothing -> do index <- fetch (m, n - 1)
                          value <- fetch (m - 1, index)
                          modify (Map.insert (m, n) value)
                          return value
            Just value -> return value
\end{minted}
\vspace{1mm}

\noindent
The main case ($m>0 \wedge n>0$) starts by looking up the pair of indices
\hs{(}\hs{m,}~\hs{n)} in the \hs{cache}. If the cache has \hs{Nothing}, we
calculate the resulting \hs{value} as before and \hs{modify} the cache
accordingly; otherwise, if we have a cache hit, we return the obtained value
immediately, skipping the actual calculation. We will use this technique in our
models of cloud build systems in~\S\ref{sec-implementations}.
